Assignment 3

### 2019-07-31 ex3/report/report.tex
last fixes

There is a significant body of scientific work focusing on the detection of unwanted behavior in networks. In the past, a viable of performing intrusion detection was to inspect the content of packets themselves and detect if a packet delivers potentially harmful content.

More recently, with the rise of omnipresent encryption, available features have changed. Since payload is not available anymore, the focus now lies on features that are always available to observers such as port numbers, flags of packets etc. Network communication is usually grouped into \textit{flows}, which is usually defined as a sequence of packets between two applications that may be located at different physical locations.

When analyzing flows, not only the aforementioned features are available but also features related to the timing of the individual packets as well as their size. For example, a specific attack might require the attacker to send three small packets and the victim to reply with two large packets followed by a small one the attacker. Various approaches have been proposed to extract features from flows and then perform anomaly detection with the results \cite{meghdouri_analysis_2018}. While this approach works well, it is problematic, that first the whole flow has to be received and only afterwards it can be determined if there was an anomaly.

@@

\begin{table}[b]
\caption{Occurence probability of attack types.} \label{tab:performance_results}
\begin{tabular}{l r} \toprule
Attack type & Proportion \\
\midrule
Normal                                                         & $0.747468$ \\
DoS / DDoS:DoS Hulk                                            & $0.101014$ \\
PortScan:PortScan - Firewall off                               & $0.069020$ \\
DDoS:LOIT                                                      & $0.040784$ \\
Infiltration:Dropbox download & $0.032982$ \\
DoS / DDoS:DoS GoldenEye                                       & $0.003224$ \\
DoS / DDoS:DoS Slowhttptest                                    & $0.001819$ \\
DoS / DDoS:DoS slowloris                                       & $0.001680$ \\
Brute Force:SSH-Patator                                        & $0.001107$ \\
Botnet:ARES                                                    & $0.000327$ \\
Web Attack:XSS                                                 & $0.000293$ \\
PortScan:PortScan - Firewall on                                & $0.000165$ \\
Brute Force:FTP-Patator                                        & $0.000110$ \\
Web Attack:Sql Injection                                       & $0.000006$ \\
DoS / DDoS:Heartbleed                                          & $0.000001$ \\
\bottomrule
\end{tabular}
\label{tab:occurence}
\end{table}

The dataset we use is the \textit{CIC-IDS 2017} dataset \cite{sharafaldin_toward_2018}, which includes the attacks listed in \autoref{tab:occurence} and contains over two million flows in total. We implement an LSTM-based classifier with three layers of 512 neurons each. As the input features we use \begin{itemize*}
\item source port
\item destination port
\item protocol identifier
\item packet sizes
\item interarrival times of packets
\item packet directions (if packet goes there or back)
\item all TCP flags (0 if flow is not TCP)
\end{itemize*}. Of these features, source port, destination port and protocol identifier are constant over the whole flow while the other vary.

@@

\autoref{tab:performance_results} shows that our classifier achieves an accuracy that is similar to the one of the random forest and the multilayer perceptron that we developed in the previous exercise. However unlike these classifier, our recurrent classifier has the advantage that it can detect attacks already before they are over.

@@

\item
\textbf{Multiple prediction outputs.}
Often times a recurrent neural network produces an output at each time step. When trying to apply explainability methods, hence, the question arises which prediction to base the methods on.

@@

\subsection{Explainability Plots}
During flow extraction we used a usual 5-tuple flow key, which distinguished flows based on the protocol they use and their source and destination port and IP address. Hence, these features are constant for all packets in one flow, which allows us to use explainability methods which have been proposed for non-recurrent classification methods.

In particular,  \autoref{fig:pdps} shows the Partial Dependency Plots (PDP) for the source and destination port features. If $\boldsymbol X \in \mathbb R ^n$ denotes a random vector drawn from feature space and $f(X) \in [0,1]$ denotes the neural network's prediction, the PDP for feature $i$ can be defined as 
\begin{equation} \label{eq:pdp}
\text{PDP}_i(y) = \mathbb E_{\boldsymbol X}\left(f(X_1,\ldots,X_{i-1},y,X_{i+1},\ldots X_n)\right)
\end{equation}

As apparent from \autoref{pdp}, for a certain value of the $i$th feature, the whole dataset is used for averaging the neural network's outcome.

@@

Next, we want to investigate if our classifier just memorized its inputs. The most interesting packet features that can be used to visualize this in our use case are the inter-arrival time (IAT) between packets in a flows and the packet length. To depict how the model comes to a decision, we determined the total minimum and maximum for these features that ever occur for all flows. For a given flow $X$ we then varied feature values at each time step between minimum and maximum and plot how the confidence changes when the feature varies (\autoref{fig:predplots2}) depict the averaged predictions for different attack families. It seems clear that the classifier didn't just memorize the inputs but actually learned them since there is always a smooth decrease of confidence if we change the feature values. Instead, when memorizing, we would expect a sharp decline of confidence to occur. Another interesting observation is that after the first couple of packets the classifier is certain about its decision and changing features doesn't change its confidence any longer. This allows the conclusion that if an attacker is able to change the first few packets, she could possibly change the decision of the classifier. We will further investigate this question in section \autoref{sec:adv}.

>>>>

There is a significant body of scientific work focusing on the detection of unwanted behavior in networks. In the past, a viable way of performing intrusion detection was to inspect the content of packets themselves and detect if a packet delivers potentially harmful content.

More recently, with the rise of omnipresent encryption, available features have changed. Since payload is not available anymore, the focus now lies on features that are always available to observers such as port numbers, flags of packets etc. Network communication is usually grouped into \textit{flows}, which is commonly defined as a sequence of packets between two applications that may be located at different physical locations.

When analyzing flows, not only the aforementioned features are available but also features related to the timing of the individual packets as well as their size. For example, a specific attack might require the attacker to send three small packets and the victim to reply with two large packets followed by a small one the attacker. Various approaches have been proposed to extract features from flows and then perform anomaly detection with the results \cite{meghdouri_analysis_2018}. While this approach works well, it is problematic that first the whole flow has to be received and only afterwards it can be determined if there was an anomaly.

@@

\begin{table}[b]
\caption{Occurence probability of attack types.} \label{tab:occurrence}
\begin{tabular}{l r} \toprule
Attack type & Proportion \\
\midrule
Normal                                                         & $0.747468$ \\
DoS / DDoS:DoS Hulk                                            & $0.101014$ \\
PortScan:PortScan - Firewall off                               & $0.069020$ \\
DDoS:LOIT                                                      & $0.040784$ \\
Infiltration:Dropbox download & $0.032982$ \\
DoS / DDoS:DoS GoldenEye                                       & $0.003224$ \\
DoS / DDoS:DoS Slowhttptest                                    & $0.001819$ \\
DoS / DDoS:DoS slowloris                                       & $0.001680$ \\
Brute Force:SSH-Patator                                        & $0.001107$ \\
Botnet:ARES                                                    & $0.000327$ \\
Web Attack:XSS                                                 & $0.000293$ \\
PortScan:PortScan - Firewall on                                & $0.000165$ \\
Brute Force:FTP-Patator                                        & $0.000110$ \\
Web Attack:Sql Injection                                       & $0.000006$ \\
DoS / DDoS:Heartbleed                                          & $0.000001$ \\
\bottomrule
\end{tabular}
\label{tab:occurence}
\end{table}

The dataset we use is the \textit{CIC-IDS 2017} dataset \cite{sharafaldin_toward_2018}, which includes the attacks listed in \autoref{tab:occurence} and contains over two million flows in total. We implement an LSTM-based classifier with three layers of 512 neurons each. As the input features we use \begin{itemize*}
\item source port
\item destination port
\item protocol identifier
\item packet sizes
\item interarrival times of packets
\item packet directions (if packet goes there or back)
\item all TCP flags (0 if flow is not TCP)
\end{itemize*}. Of these features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.

@@

\autoref{tab:performance_results} shows that our classifier achieves an accuracy that is similar to the one of the random forest and the multilayer perceptron that we developed in the previous exercise. However unlike these classifiers, our recurrent classifier has the advantage that it can detect attacks already before they are over.

@@

\item
\textbf{Multiple prediction outputs.}
Often a recurrent neural network produces an output at each time step. When trying to apply explainability methods, hence, the question arises which prediction to base the methods on.

@@

\subsection{Explainability Plots}
During flow extraction we used a usual 5-tuple flow key, which distinguishe flows based on the protocol they use and their source and destination port and IP address. Hence, these features are constant for all packets in one flow, which allows us to use explainability methods which have been proposed for non-recurrent classification methods.

In particular,  \autoref{fig:pdp} shows the Partial Dependency Plots (PDP) for the source and destination port features. If $\boldsymbol X \in \mathbb R ^n$ denotes a random vector drawn from feature space and $f(X) \in [0,1]$ denotes the neural network's prediction, the PDP for feature $i$ can be defined as 
\begin{equation} \label{eq:pdp}
\text{PDP}_i(y) = \mathbb E_{\boldsymbol X}\left(f(X_1,\ldots,X_{i-1},y,X_{i+1},\ldots X_n)\right)
\end{equation}

As apparent from \autoref{fig:pdp}, for a certain value of the $i$th feature, the whole dataset is used for averaging the neural network's outcome.

@@

Next, we want to investigate if our classifier just memorized its inputs. The most interesting packet features that can be used to visualize this in our use case are the inter-arrival time (IAT) between packets in a flows and the packet length. To depict how the model comes to a decision, we determined the total minimum and maximum for these features that ever occur for all flows. For a given flow $X$ we then varied feature values at each time step between minimum and maximum and plot how the confidence changes when the feature varies (\autoref{fig:pred_plots2}) depict the averaged predictions for different attack families. It seems clear that the classifier didn't just memorize the inputs but actually learned them since there is always a smooth decrease of confidence if we change the feature values. Instead, when memorizing, we would expect a sharp decline of confidence to occur. Another interesting observation is that after the first couple of packets the classifier is certain about its decision and changing features doesn't change its confidence any longer. This allows the conclusion that if an attacker is able to change the first few packets, she could possibly change the decision of the classifier. We will further investigate this question in section \autoref{sec:adv}.
